\texttt{aeons}: approximating the end of nested sampling

### 2023-11-29 paper/paper.tex
Adjustments made

A constant step size in $\log X$ corresponds to a geometrically constant measure for the dead points, which is exactly needed to overcome the curse of dimensionality.

@@

\subsection{Log-likelihood}\label{sec:logL}
We now consider the distribution of the samples in log-likelihood. To get an insight into the analytics, we will examine the representative case of the $d$-dimensional multivariate Gaussian with maximum point $\logLmax$ and length scale $\sigma$:
\begin{equation}\label{eq:gaussian_logL}
    \log \Like = \logLmax - X^{2/d}/2\sigma^2.
\end{equation}
Providing that $\sigma^d \ll 1$, the posterior can be approximated as 
\begin{equation}
    \mathcal{P}(X) = \tfrac{1}{(2\pi\sigma^2)^{d/2}} e^{-X^{2/d}/2\sigma^2}, 
    \label{eqn:PX}
\end{equation}
which in terms of $\log\mathcal{L}$ becomes
\begin{equation}
    \mathcal{P}(\log\mathcal{L}) = \frac{1}{\Gamma\left(\frac{d}{2}\right)}e^{\log\mathcal{L}-\log\mathcal{L}_\mathrm{max}} (\log\mathcal{L}_\mathrm{max}-\log\mathcal{L})^{\frac{d}{2}-1}
\end{equation}
i.e. $2(\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}) \sim \chi^2_{d}$, with mean and variance
\begin{equation}
    \langle\log\mathcal{L}\rangle_\mathcal{P} = \log\mathcal{L}_\mathrm{max} - \frac{d}{2},  \quad \mathrm{Var}(\log\mathcal{L})_\mathcal{P} = \frac{d}{2}.
\end{equation}
Given the prior is uniform $X\sim U(0,1)$, the KL divergence in the $\sigma^d \ll 1$ limit is:
\begin{equation}
\mathcal{D}_\mathrm{KL} = \frac{d}{2}\log 2 \Gamma(1+\tfrac{d}{2})^{2/d} \sigma^2
\end{equation}

@@

    P(\log\Like) = \frac{\mathrm{d} \log X}{\mathrm{d} \log\Like} P(\log X) \propto \frac{dn}{2(\logLmax - \log\Like)},

@@

How much further do the live points penetrate in log-likelihood? We seek the distribution for the maximum likelihood of the live points, $\log\Like_{(n_i)}^{\mathrm{live}}$, where we have used the notation of order statistics to denote $x_{(n)}$ as the maximum of $n$ points.  For convenience, we will normalise the likelihoods as

@@

or in the large $d$, $n$ limit

@@

    \lim_{d\to\infty} y_{(n_i)}^\mathrm{live} &\sim \frac{2H_n}{d} \pm \left(\frac{2(\pi^2 - 6\Psi^{(1)}(1+n))}{3d^2}\right)^{\frac{1}{2}},

@@

    \lim_{d,n\to\infty} y_{(n_i)}^\mathrm{live} &\sim \frac{2\log n}{d} \pm \sqrt{\frac{2}{3}}\frac{\pi}{d},

@@

where $\psi^{(1)}$ is the trigamma function and $H_n$ is the $n$th harmonic number. This is a very small fraction in high dimensions, showing that until the end the live points are far from the posterior bulk. 

@@

        \log{\mathcal{L}}_\mathrm{max}^\mathrm{live} \approx \log\mathcal{L}_\mathrm{max} - \frac{d}{2e} + \frac{\log n}{e} \pm \frac{\pi}{\sqrt{6}e}

@@

showing that in general nested sampling will finish at a contour $d/2e$ away from the maximum log-likelihood. The final set of $n$ live points gets you $\log n/e$ closer, with a chance of getting $\sim\pi/\sqrt{6}e=0.472$ closer still by statistical fluctuation. 

@@

Making the termination criterion stricter therefore has limited returns in high-dimensions, if it is ultimately still based on the remaining evidence. However, nested sampling still shrinks around the maximum exponentially, so provided a good termination criterion is chosen, it will get there in reasonable time.

To quantify this statement, consider the number of iterations required to get from the posterior bulk to the true maximum, which we will now calculate. At the posterior, we have roughly speaking
\begin{equation}
    (\log X, \log\Like) = (-\DKL, \logLmax - d/2).
\end{equation}
The prior volume $\log X_\delta$ at which the likelihood is within $\delta$ of the maximum can then be found by inverting \cref{eq:gaussian_logL}:
\begin{equation}
    \log X_\delta = -\DKL - \frac{d}{2} \left(\log \frac{d}{2} - \log \delta\right),
\end{equation}
which corresponds to an additional
\begin{equation}
    \Delta i = \frac{nd}{2} \left(\log \frac{d}{2} - \log \delta\right)
\end{equation}
iterations after the bulk is reached. One would therefore expect a general distribution to take $\mathcal{O}(\tfrac{n d }{2}\log \tfrac{d}{2})$ iterations to get from the usual nested sampling stopping point to within an $e$-fold of the maximum, which can serve as a sanity check for the maximisation procedure. 
\par
A summary of the distances between the notable points at the end of a run is shown in \cref{fig:logL_distribution}.

@@

\section{Endpoint prediction}\label{sec:endpoint}
As described in \citet{supernest} and further explored in the talk and upcoming paper \citet{kcl_talk, scaling_frontier}, the time complexity of nested sampling is
\begin{equation}
    T \propto \langle \tfrac{1}{\nlive} \rangle^{-1} \times \langle \mathcal{T}\{ \Like(\theta) \} \rangle \times \langle f_\mathrm{sampler} \rangle \times \DKL.
\end{equation}
The first term is the harmonic mean of the number of live points $\sim\mathcal{O}(n)$, The second term is the average time per likelihood evaluation. The third is the average number of evaluations required to replace a dead point with a live point at higher likelihood, which is given by the implementation and usually does not vary in orders of magnitude. 
The final term is the Kullback-Liebler divergence~\cref{eq:DKL}, the compression factor required to get from the prior to the posterior. This term is generally outside of user control, in most cases \textit{a priori} unknown, and of principle interest in this section.

>>>>

A steady step size in $\log X$ corresponds to a geometrically constant measure for the dead points, which is exactly needed to overcome the curse of dimensionality.

@@

\subsection{Log-likelihood}\label{sec:logL}
We now consider the distribution of the samples in log-likelihood. To get an insight into the analytics, we will examine the representative case of the $d$-dimensional multivariate Gaussian with maximum point $\logLmax$ and length scale $\sigma$:
\begin{equation}\label{eq:gaussian_logL}
    \log \Like = \logLmax - X^{2/d}/2\sigma^2.
\end{equation}
Providing that $\sigma^d \ll 1$, the posterior can be approximated as 
\begin{equation}
    \mathcal{P}(X) = \tfrac{1}{(2\pi\sigma^2)^{d/2}} e^{-X^{2/d}/2\sigma^2}, 
    \label{eqn:PX}
\end{equation}
which in terms of $\log\mathcal{L}$ becomes
\begin{equation}
    \mathcal{P}(\log\mathcal{L}) = \frac{1}{\Gamma\left(\frac{d}{2}\right)}e^{\log\mathcal{L}-\log\mathcal{L}_\mathrm{max}} (\log\mathcal{L}_\mathrm{max}-\log\mathcal{L})^{\frac{d}{2}-1}
\end{equation}
i.e. $2(\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}) \sim \chi^2_{d}$, with mean and variance
\begin{equation}
    \langle\log\mathcal{L}\rangle_\mathcal{P} = \log\mathcal{L}_\mathrm{max} - \frac{d}{2},  \quad \mathrm{Var}(\log\mathcal{L})_\mathcal{P} = \frac{d}{2}.
\end{equation}
Given the prior is uniform $X\sim U(0,1)$, the KL divergence in the $\sigma^d \ll 1$ limit is:
\begin{equation}
\mathcal{D}_\mathrm{KL} = -\frac{d}{2}\log 2 \Gamma(1+\tfrac{d}{2})^{2/d} \sigma^2
\end{equation}

@@

    P(\log\Like) = \frac{\mathrm{d} \log X}{\mathrm{d} \log\Like} P(\log X) \propto \frac{d\:n_i}{2(\logLmax - \log\Like)},

@@

How much further do the live points penetrate in log-likelihood? We seek the distribution for the maximum likelihood of the live points, $\log\Like_{(n_i)}^{\mathrm{live}}$, where we have used the notation of order statistics to denote $x_{(k)}$ as the maximum of $k$ points.  For convenience, we will normalise the likelihoods as

@@

or in the large $d$, $n_i$ limit

@@

    \lim_{d\to\infty} y_{(n_i)}^\mathrm{live} &\sim \frac{2H_{n_i}}{d} \pm \left(\frac{2(\pi^2 - 6\Psi^{(1)}(1+n_i))}{3d^2}\right)^{\frac{1}{2}},

@@

    \lim_{d,n_i\to\infty} y_{(n_i)}^\mathrm{live} &\sim \frac{2\log n_i}{d} \pm \sqrt{\frac{2}{3}}\frac{\pi}{d},

@@

where $\psi^{(1)}$ is the trigamma function and $H_{n_i}$ is the $n_i$th harmonic number. This is a very small fraction in high dimensions, showing that until the end the live points are far from the posterior bulk. 

@@

        \log{\mathcal{L}}_\mathrm{max}^\mathrm{live} \approx \log\mathcal{L}_\mathrm{max} - \frac{d}{2e} + \frac{\log n_i}{e} \pm \frac{\pi}{\sqrt{6}e}

@@

showing that in general nested sampling will finish at a contour $d/2e$ away from the maximum log-likelihood. The final set of $n_i$ live points gets you $\log n_i/e$ closer, with a chance of getting $\sim\pi/\sqrt{6}e=0.472$ closer still by statistical fluctuation. 

@@

Making the traditional termination criterion stricter therefore has limited returns in high-dimensions, if it is ultimately still based on the remaining evidence. However, nested sampling still shrinks around the maximum exponentially, so provided a good alternative termination criterion is chosen, it will get there in reasonable time.

To quantify this statement, consider the number of iterations $\Delta i$ required to get from the posterior bulk to the true maximum, which we will now calculate. At the posterior, we have roughly speaking
\begin{equation}
    (\log X, \log\Like) = (-\DKL, \logLmax - d/2).
\end{equation}
The prior volume $\log X_\delta$ at which the likelihood is within $\delta$ of the maximum can then be found by inverting \cref{eq:gaussian_logL}:
\begin{equation}
    \log X_\delta = -\DKL - \frac{d}{2} \left(\log \frac{d}{2} - \log \delta\right),
\end{equation}
which corresponds to an additional
\begin{equation}
    \Delta i = \frac{nd}{2} \left(\log \frac{d}{2} - \log \delta\right)
\end{equation}
iterations after the bulk is reached, where $n$ is the harmonic mean of the number of live points. One would therefore expect a general distribution to take $\mathcal{O}(\tfrac{n d }{2}\log \tfrac{d}{2})$ iterations to get from the usual nested sampling stopping point to within an $e$-fold of the maximum. A rule-of-thumb termination criterion could therefore be to run for at least $\tfrac{nd}{2}\log\tfrac{d}{2}$ iterations after the posterior is reached.
\par
A summary of the distances between the notable points at the end of a run is shown in \cref{fig:logL_distribution}.

@@

\section{Endpoint prediction}\label{sec:endpoint}
As described in \citet{supernest} and further explored in the talk and upcoming paper \citet{kcl_talk, scaling_frontier}, the time complexity of nested sampling is
\begin{equation}
    T \propto \langle \tfrac{1}{\nlive} \rangle^{-1} \times \langle \mathcal{T}\{ \Like(\theta) \} \rangle \times \langle f_\mathrm{sampler} \rangle \times \DKL.
\end{equation}
The first term is the harmonic mean of the number of live points $\sim\mathcal{O}(n_i)$, The second term is the average time per likelihood evaluation. The third is the average number of evaluations required to replace a dead point with a live point at higher likelihood, which is given by the implementation and usually does not vary in orders of magnitude. 
The final term is the Kullback-Liebler divergence~\cref{eq:DKL}, the compression factor required to get from the prior to the posterior. This term is generally outside of user control, in most cases \textit{a priori} unknown, and of principle interest in this section.
